Reformulating phylogenetic mixed models to improve flexibility and speed

### 2022-08-11 phyloglmm_ms_resubmit.tex
misc TODO/fix multi-time plot
MLi tweaks and fixing some technical problems

\bmb{ML: please check notation; in particular, is $\sigma^2_{\mathrm{phy}}$ correct??}
Compared to the single--group model, the multi-group simulation model has five additional random effect parameters: predictor-level ($\sigma^2_{\mathrm{sp_{slope}}}$) and intercept-level ($\sigma^2_{\mathrm{sp_{int}}}$) among-species variances;  their covariance ($\sigma_{\mathrm{sp_{int-slope}}}$);  among-group variation in the intercept ($\sigma^2_{\mathrm{group}}$); and variation among species-group combinations in the intercept ($\sigma^2_{\mathrm{phy}}$).
Because each species has multiple observations we can distinguish variation among species from residual variation, and thus we can include predictor-level and intercept-level random effects of species.
Variance in the intercept of species--group interactions ($\sigma^2_{\mathrm{phy}}$) describes whether the species within a group have more similar responses on average than expected by chance, equivalent to phylogenetic attraction \citep{helmus2007separating}.

@@

Using the R package \pkg{ape} \citep{ape}, we simulated 100 random phylogenetic trees  for each sample size ($n$ = 25, 50, 100 and an additional $n$ = 500 for the multi-group model) and then simulated the responses for each tree (\ref{eq:single_glmm}, \ref{eq:multiple_glmm}). 
Each realization was fitted using all model variants. 
All simulation parameters are shown in Figure \ref{ssplot} and Figure \ref{msplot}. 
Table \ref{table:platform} shows the parameters that are estimable for each platform. 
We only evaluated the goodness of fit for model fits that passed the convergence tests implemented by the package.
For Bayesian fits, we evaluate realizations with Gelman-Rubin statistic $<1.1$. Based on recent concerns about Gelman-Rubin thresholds \citep{vats2021revisiting}, we additionally restricted results to fits with  effective sample size $>1000$ for the fixed effect parameters  \citep[$\beta_{0}$ and $\beta_{1}$: ][]{vehtari2021rank}. 
For each replicate, we sample two chains starting with 10000 iterations. 
We first evaluate our estimates by looking at the distribution of the estimated values (maximum likelihood estimates for non-Bayesian platforms and posterior medians for Bayesian platforms) to quantify bias and variance (i.e., quality of the point estimate).
We computed 95\% Wald confidence intervals for frequentist methods \bmb{check: I would think lme4 would only give profile CIs for RE parameters?} and quantile-based intervals for Bayesian methods, then computed coverage --- the proportion of simulations in which the computed confidence intervals include the true values of parameters --- to assess the quality of the confidence intervals.
We also compare computational speed between different platforms.

@@

\begin{center}
\begin{figure}[H]
  \includegraphics[scale=0.7]{./figure/ssplot.pdf}
  \caption{Comparison of single group model parameter estimates across different R packages in Table \ref{table:platform}.
    Total simulations $n=100$ for each category. The horizontal line shows the true value of the parameters in the simulation model. Models capable of fitting all parameters (\pkg{\plmefour},  \pkg{\pgTMB}, and \pkg{brms}) fit well for all parameters.
    \bmb{are we setting a boundary at 200? why are some methods pegging out here?}
%\bmb{consider tweaking aspect ratio (slightly wider?) Consider including n=25, n=50, n=100 in x-axis tick labels (I don't have all the output data so can't rebuild figs myself ...)
%\jd{Use dashed lines (or some other gg signifier) for the insufficient approaches}
}
\label{ssplot}
\end{figure}
\end{center}
\begin{center}
\begin{figure}[H]
  \includegraphics[scale=0.7]{./figure/csstime.pdf}
  \caption{Single-group model computational speed. Dashed lines indicate scaling relationships: lines parallel to the lower line would indicate linear scaling of computational time with number of species,
  while those parallel to the upper line would indicate quadratic scaling.}
\label{ssplot_speed}
\end{figure}
\end{center}

@@

\begin{center}
\begin{figure}[H]
  \includegraphics[scale=0.7]{./figure/msplot.pdf}
  \caption{Multi-group model parameter estimates. The horizontal line shows the true value of the parameters in the simulation model. Models capable of fitting all parameters (\pkg{\plmefour} and \pkg{\pgTMB}) fit well for all parameters. \pkg{pez} and \pkg{phyr} estimates for $n$ = 500 are missing because the models did not converge within 30 minutes.
  }
  \label{msplot}
\end{figure}
\end{center}
\begin{center}
\begin{figure}[H]
  \includegraphics[scale=0.7]{./figure/mstime.pdf}
  \caption{Multi-group model computational speed.}
  \label{msplot_time}
\end{figure}
\end{center}
\begin{center}
\begin{figure}[H]
  \includegraphics[scale=0.7]{./figure/mscoverage.pdf}
  \caption{Comparison of multi-group model coverage. The black line shows the nominal coverage, and the blue ribbon the 95\% binomial confidence interval based on 100 simulated fits.}
  \label{msplot_coverage}
\end{figure}
\end{center}

>>>>

Compared to the single--group model, the multi-group simulation model has five additional random effect parameters: predictor-level ($\sigma^2_{\mathrm{sp_{slope}}}$) and intercept-level ($\sigma^2_{\mathrm{sp_{int}}}$) among-species variances;  their covariance ($\sigma_{\mathrm{sp_{int-slope}}}$);  among-group variation in the intercept ($\sigma^2_{\mathrm{group}}$); and variation among species-group combinations in the intercept ($\sigma^2_{\mathrm{phy}}$).
Because each species has multiple observations we can distinguish variation among species from residual variation, and thus we can include predictor-level and intercept-level random effects of species.
Variance in the intercept of species--group interactions ($\sigma^2_{\mathrm{phy}}$) describes whether the species within a group have more similar responses on average than expected by chance, equivalent to phylogenetic attraction \citep{helmus2007separating}.

@@

Using the R package \pkg{ape} \citep{ape}, we simulated 100 random phylogenetic trees  for each sample size ($n$ = 25, 50, 100 and an additional $n$ = 500 for the multi-group model) and then simulated the responses for each tree (\ref{eq:single_glmm}, \ref{eq:multiple_glmm}). 
Each realization was fitted using all model variants. 
All simulation parameters are shown in Figure \ref{ssplot} and Figure \ref{msplot}. 
Table \ref{table:platform} shows the parameters that are estimable for each platform. 
We only evaluated the goodness of fit for model fits that passed the convergence tests implemented by the package.
For Bayesian fits, we evaluate realizations with Gelman-Rubin statistic $<1.1$. Based on recent concerns about Gelman-Rubin thresholds \citep{vats2021revisiting}, we additionally restricted results to fits with  effective sample size $>1000$ for the fixed effect parameters  \citep[$\beta_{0}$ and $\beta_{1}$: ][]{vehtari2021rank}. 
For each replicate, we sample two chains starting with 10000 iterations. 
We first evaluate our estimates by looking at the distribution of the estimated values (maximum likelihood estimates for non-Bayesian platforms and posterior medians for Bayesian platforms) to quantify bias and variance (i.e., quality of the point estimate).
We computed 95\% Wald confidence intervals for frequentist methods and quantile-based intervals for Bayesian methods, then computed coverage --- the proportion of simulations in which the computed confidence intervals include the true values of parameters --- to assess the quality of the confidence intervals.
We also compare computational speed between different platforms.

@@

\begin{center}
\begin{figure}[H]
  \includegraphics[scale=0.7]{./figure/ssplot.pdf}
  \caption{Comparison of single group model parameter estimates across different R packages in Table \ref{table:platform}.
    Total simulations $n=100$ for each category. The horizontal line shows the true value of the parameters in the simulation model. Models capable of fitting all parameters (\pkg{\plmefour},  \pkg{\pgTMB}, and \pkg{brms}) fit well for all parameters. Values above 200 (very common for \pkg{gls} and \pkg{phylolm}, less than 1\% of results otherwise) are censored to 200.
    %\bmb{consider tweaking aspect ratio (slightly wider?) Consider including n=25, n=50, n=100 in x-axis tick labels (I don't have all the output data so can't rebuild figs myself ...)
%\jd{Use dashed lines (or some other gg signifier) for the insufficient approaches}
}
\label{ssplot}
\end{figure}
\end{center}
\begin{center}
\begin{figure}[H]
  \includegraphics[scale=0.7]{./figure/csstime.pdf}
  \caption{Single-group model computational speed. Dashed lines indicate scaling relationships: lines parallel to the lower line would indicate linear scaling of computational time with number of species,
  while those parallel to the upper line would indicate quadratic scaling.}
\label{ssplot_speed}
\end{figure}
\end{center}

@@

\begin{center}
\begin{figure}[H]
  \includegraphics[scale=0.7]{./figure/msplot.pdf}
  \caption{Multi-group model parameter estimates. The horizontal line shows the true value of the parameters in the simulation model. Models capable of fitting all parameters (\pkg{\plmefour} and \pkg{\pgTMB}) fit well for all parameters. \pkg{pez} and \pkg{phyr} estimates for $n$ = 500 are missing because the models did not converge within 30 minutes.
  }
  \label{msplot}
\end{figure}
\end{center}
\begin{center}
\begin{figure}[H]
  \includegraphics[scale=0.7]{./figure/mstime.pdf}
  \caption{Multi-group model computational speed. Linear and quadratic scaling lines as in Figure~\ref{ssplot_speed}.}
  \label{msplot_speed}
\end{figure}
\end{center}
\begin{center}
\begin{figure}[H]
  \includegraphics[scale=0.7]{./figure/mscoverage.pdf}
  \caption{Comparison of multi-group model coverage. The black line shows the nominal coverage, and the blue ribbon the 95\% binomial confidence interval based on 100 simulated fits.}
  \label{msplot_coverage}
\end{figure}
\end{center}
